Combining information from photometric and spectroscopic data:\\ Don't waste your time in spectrophotometric calibration!

### 2014-04-27 tex/ms.tex
more ex
bdj changes to model generalities

% ## style notes:
% - all equations in `eqnarray` environments
% - parameters are `inferred`; `measurement` is reserved for data
% - use emulateapj or aastex, as you like, but watch for equation-wrapping in the former

% ---- Hogg (aastex) mode ----
% \documentclass[12pt, letterpaper, preprint]{aastex}
% ----

% ---- Charlie (emulateapj) mode ----
\documentclass[iop,numberedappendix]{emulateapj}
\usepackage{apjfonts}
% ----

@@

% math shih
\newcommand{\transpose}[1]{{#1}^{\!\mathsf T}}
\newcommand{\given}{\,|\,}

@@

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - f(\alpha)\ast\mu(\theta)
\\
C_{nn'} &\equiv& \sigma_n^2\,\delta_{nn'} + K_\alpha(\lambda_n - \lambda_{n'})
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
the ``$\ast$'' operator is the pixel-wise multiplication operator,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.

In detail, we are going to choose a $M$th-order polynomial form for
the calibration vector $f(\alpha)$ and a bell shape (Gaussian) for the
kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}
f(\alpha) &=& \sum_{m=1}^M c_m\,x^m
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(x) &=& a^2\,\exp(-\frac{x^2}{2\,\ell^2})
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths $\lambda_m$
and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise.

@@

>>>>

% ## style notes:
% - all equations in `eqnarray` environments
% - use `\,` as the multiply operator!
% - parameters are `inferred`; `measurement` is reserved for data
% - use emulateapj or aastex, as you like, but watch for equation-wrapping in the former

% ---- Hogg (aastex) mode ----
\documentclass[12pt, letterpaper, preprint]{aastex}
% ----

% ---- Charlie (emulateapj) mode ----
% \documentclass[iop,numberedappendix]{emulateapj}
% \usepackage{apjfonts}
% ----

@@

% math shih
\newcommand{\transpose}[1]{{#1}^{\!\mathsf T}}
\newcommand{\given}{\,|\,}
\renewcommand{\det}[1]{||{#1}||}

@@

If you don't believe the reported data noise variances $\sigma_n^2$,
it often makes sense to add in quadrature a ``jitter'' variance $s^2$.
This is a trivial modification of the chi-squared log-likelihood:
\begin{eqnarray}\label{eq:photometricLF}
\ln p(d\given\theta,s^2) &=& -\frac{1}{2}\,\sum_{n=1}^N \left[\frac{[d_n - \mu_n(\theta)]^2}{[\sigma_n^2 + s^2]} + \ln(2\pi\,[\sigma_n^2 + s^2]) \right]
\quad .
\end{eqnarray}
This modification shows the necessity of including the log-variance
term.
This---equation~(\ref{eq:photometricLF})---is in fact the log-likelihood
function we are going to use for the photometric measurements (BDJ: IS
THIS TRUE?) in what follows.
In this formulation, if $\theta$ are the parameters of interest, the
jitter variance $s^2$ is a nuisance noise-model parameter.
The Gaussian Process will be a generalization of this modification.

The calibration-vector and Gaussian-Process modified log-likelihood is
\begin{eqnarray}\label{eq:spectroscopicLF}
\ln p(d\given\theta,\alpha) &=& -\frac{1}{2}\,\left[\transpose{\Delta}\,C^{-1}\,\Delta + \ln([2\pi]^N\,\det{C}) \right]
\\
\Delta &\equiv& d - f(\alpha)\ast\mu(\theta)
\\
C_{nn'} &\equiv& [\sigma_n^2 + s^2]\,\delta_{nn'} + K_\alpha(\lambda_n - \lambda_{n'})
\quad ,
\end{eqnarray}
where $\alpha$ is a blob of nuisance parameters
(which determine the jitter, the calibration vector and the kernel function),
$\Delta$ is a residual column vector,
$f(\alpha)$ is a calibration column vector,
the ``$\ast$'' operator is the pixel-wise multiplication operator,
$C$ is the Gaussian Process covariance matrix,
made up of $N^2$ values $C_{nn'}$,
$K_\alpha$ is a kernel function,
and each $\lambda_n$ is the wavelength of spectral pixel $n$.
The calibration vector $f(\alpha)$ is going to take care of the
large-scale calibration residual; the Gaussian process is going to
pick up smaller-scale residuals away from \emph{that}.
This---equation~(\ref{eq:spectroscopicLF})---is the log-likelihood we
are going to use for the spectroscopic data in what follows.

In detail, we are going to choose a $M$th-order polynomial form for
the calibration vector $f(\alpha)$ and a bell shape (Gaussian) for the
kernel function $K_\alpha(\cdot)$:
\begin{eqnarray}\displaystyle
f(\alpha) &=& c_0\,[1 + \sum_{m=1}^M c_m\,x^m]
\\
x &\equiv& \frac{\lambda}{\lambda_0} - 1
\\
K_\alpha(\lambda_n - \lambda_{n'}) &=& a^2\,\exp(-\frac{[\lambda_n - \lambda_{n'}]^2}{2\,\ell^2})
\\
\transpose{\alpha} &\equiv& \left[ \{c_m\}_{m=0}^M, s^2, a^2, \ell^2 \right]
\quad ,
\end{eqnarray}
where $\lambda$ is a column vector made up of the wavelengths $\lambda_m$
and $x$ is a scaled and shifted version of the same,
scaled by a fiducial wavelength $\lambda_0$.
Nuisance parameters $c_m$ set the shape of the large-scale calibration
issues, while $a^2$ and $\ell^2$ set the variance and correlation
length of the remaining covariant calibration noise.

@@

We are now left to demonstrate that it is possible to infer the
parameters of interest $\theta$ while marginalizing over the
calibration parameters $\alpha$.
